\texttt{treats}: a modular \texttt{R} package for simulating trees and traits.

### 2023-06-19 paper/treats_paper.tex
fixed grants
computer reboot update

\title{\texttt{treats}: a modular \texttt{R} package for simulating
trees and traits.}
\author{Thomas Guillerme\(^1\)}
\date{2023-06-07}

@@

Comparing biological patterns is one of the key ways to understand
mechanisms in evolutionary biology. This lead to the development of
phylogenetic comparative methods as key methodologically driven topic in
ecology, evolution and palaeontology (Felsenstein 1985; Pennell and
Harmon 2013). As indicated in the name, phylogenetic comparative methods
rely on comparing patterns in a phylogenetic context to understand
biological mechanisms or concepts (Harmon 2019). These comparisons can
be done between observed patterns under different conditions or against
null, neutral or baseline models (see Bausman (2018) for distinctions)
suggesting different processes or mechanisms. For example different
traits distribution for species with different diets (Deepak, Gower, and
Cooper 2023), or habitats (Pinto-Ledezma et al. 2017) or by comparing
some observed pattern to one simulated under null or base conditions
(Miller et al. 2022). In theory, workers can follow the research
pipeline of thinking of a specific mechanism (e.g.~mass extinction
allowing the surviving species to acquire new morphologies), collect
some data to test this mechanism (e.g.~some traits of species across and
extinction event) and then compare these patterns to one simulated under
no specific conditions (e.g.~a null model where the traits evolve
randomly regardless of an extinction event Puttick, Guillerme, and Wills
(2020)). For such an approach, we need statistical and softwares
solutions to simulate trees and data to generate these specific null
models.

In practice, these evolutionary simulations can be done relatively
easily on computers using a birth-death process (Feller 1939; Stadler
2010; FitzJohn 2012). A birth-death process is a continuous time Markov
process that had been routinely implemented in \texttt{R} (R Core Team
2023) to simulate biological phylogenies (e.g.~(Paradis and Schliep
2019), (FitzJohn 2012)). These algorithm to generate phylogenetic trees
can be coupled with other Markov processes to also generate traits, for
example using a Brownian Motion process (BM; Cavalli-Sforza and Edwards
(1967)) or an Ornstein Uhlenbeck (OU; Lande (1976); see Cooper et al.
(2016) for a distinction between both). In \texttt{R}, this can be done
with several already well used and well documented packages. For example
if you want to simulate diversity through time, you can use
\texttt{TreeSim} (Stadler 2011) to simulate diversity under a set of
specific parameters (e.g.~speciation and extinction) with some events
disrupting the simulations (e.g.~mass extinctions). One can even improve
on generating these patterns using \texttt{FossilSim} (Barido-Sottani et
al. 2019) to generate a pattern that will take into account
fossilization processes. Or use \texttt{paleobuddy} (Rosario Petrucci,
Januario, and Quental 2022) or \texttt{paleotree} (Bapst 2012) to
generate palaeontology specific data. On the other hand, if you need to
simulate both diversity and traits through time, this can be done with
specific parameters in \texttt{RPANDA} (Morlon et al. 2016) or in
\texttt{PETER} (Puttick, Guillerme, and Wills 2020) where the traits are
generated stochasticaly through time (given some process) during the
birth-death process.

Although the packages mentioned above are excellent and routinely used
with fast and reliable algorithms and associated documentation, they are
all designed for specific tasks and don't allow much modification beyond
the input parameters designed by the authors. For example,
\texttt{TreeSim} can simulate a birth-death tree with some extinction
event but is not designed to simulated one with an extinction event that
leads to the birth-death process to be not diversity dependent anymore
(e.g.~simulating a release in selection pressure after the extinction
process that leads to a different process dominating speciation). Or
\texttt{PETER} is not designed to simulate a complex set of traits (say
three correlated BM traits and two independent OU ones). This absence of
modularity has hampered the use of complex and question tailored
simulations evolutionary biology, although I acknowledge this was not
the primary aim of the authors of the packages mentioned above. This has
led workers to often develop their own tools to answer specific
questions (e.g.~Puttick, Guillerme, and Wills (2020)). Therefore, we
propose \texttt{treats} a modular \texttt{R} package to simulate both
diversity and disparity through time.

@@

These four steps are repeated until the tree reaches the desired age or
number of species. If traits are simulated during the process, a fourth
step is added:

@@

In \texttt{treats}, these three or four steps are implemented as modular
functions that the user can easily change using an internal class of
objects called \texttt{"modifiers"} or \texttt{"traits"} (Fig.
\ref{Fig:workflow}). The simulation then outputs a tree (of class
\texttt{"phylo"} and a associated table of traits - \texttt{"matrix"})
that can be visualised using the \texttt{plot.treats} function. A third
class of object called \texttt{"events"} can be added to the simulations
to modify it under certain conditions (e.g.~simulating a mass
extinction). Each element in the algorithm described above can be
modified by the user using the implemented functions
\texttt{make.bd.params} to set the birth-death parameters,
\texttt{make.traits} to set the trait(s), \texttt{make.modifiers} to set
the birth-death algorithm and \texttt{make.events} to eventually add one
or more events.

@@

The modularity structure of the \texttt{treats} packages allows users to
design complex simulation scenarios to fit their specific question and
data. To illustrate this we will look whether it is possible to detect
changes in disparity (i.e.~diversity of traits) in a subset of the data
published from Beck and Lee (2014) implemented in Guillerme (2018). This
dataset contains the ordinated traits for 50 mammalian species across
the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).

@@

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
\caption{Figure 2: Observed phylogeny (A) and disparity through time (B)
in the subset of Beck and Lee (2014)'s dataset. For this specific data,
is it possible to detect potential changes of disparity due to a mass
extinction event?}
\end{figure}

@@

The first and simplest way is to simulate tree topologies that have
similar properties than the observed one. To do so, we need to use some
\textbf{speciation} parameter indicating the rate at which lineages
speciate (\emph{aka} ``birth'' or ``\(\lambda\)'') and an
\textbf{extinction} parameter indicating the rate at which they go
extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
be crudely calculated from a give tree by counting the number of
speciations and extinctions through time (but see better methods,
e.g.~\texttt{ape::birthdeath}; Lawton, May, and Raup (1995); Paradis and
Schliep (2019)). We also need a stopping rule for when to stop the
simulations (in our case when reaching 30 time units). This will produce
a single random tree using the input parameters (Fig. 3). Note that I
will not discuss the options here in great details. Much more
information can be found in the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.

@@

To answer our question, we also want to simulate an extinction event. To
do so, we can create two different \texttt{"events"} object with the
function \texttt{make.events}. The first one will simulate a random
extinction event after reaching half of the simulations (15 time units)
and then making three quarter (0.75) of the taxa go extinct. The second
one will simulate a random extinction but based on trait values: after
reaching time 20, all the species with positive trait values will go
extinct. Both scenarios thus illustrate two different types of mass
extinction (selective and random Puttick, Guillerme, and Wills (2020)).

@@

From these results we can draw some preliminary conclusions: the
observed change in disparity is more likely due to a random mass
extinction rather than a selective one. This is of course a crude way of
testing this, a more rigorous approach is likely needed to answer the
question (i.e.~more and better quality data, and more thorough methods,
e.g.~using a rank envelope test Murrell (2018)).

@@

The following sections provide an overview of the three main functions
displayed in Figure 1 (\texttt{make.traits}, \texttt{make.modifiers} and
\texttt{make.events}).

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Functionalising the previous example of a Brownian Motion process}
\NormalTok{my.bm.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, edge.length) \{}
    \KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ x0, }\DataTypeTok{sd =}\NormalTok{ edge.length)}
\NormalTok{\} }
\CommentTok{\#\# Creating the traits object}
\NormalTok{my.trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ my.bm.process)}
\end{Highlighting}
\end{Shaded}

Of course biological data can be much more complex and often
multivariate (Adams and Collyer 2019) requiring the users to develop
more complex function to cater their specific needs. However, the
\texttt{treats} package contains a list of pre-built processes that are
``ready-to-use'':

\begin{itemize}
\tightlist
\item
  \texttt{BM.process} and \texttt{OU.process}: generalised Brownian
  Motion and Ornstein-Uhlenbeck processes in any number of dimensions
  (including possible correlation);
\item
  \texttt{multi.peak.process}: a generalised Ornstein-Uhlenbeck process
  (uni- or multi-dimensional) which allows for multiple peaks (usually
  called the ``alpha'' parameters; Lande (1976));
\item
  \texttt{repulsion.process}: a unidimensional process generating trait
  values that don't overlap with previously generated trait values;
\item
  \texttt{discrete.process}: generates discrete trait values;
\item
  \texttt{no.process}: ignores the branch length (i.e.~a time
  independent process);
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  C\textbf{hoosing whether that lineage goes extinct (becomes a tip) or
  speciates (becomes a node)} by drawing a random number between 0 and 1
  and comparing it to the ratio of speciation and turnover (speciation +
  extinction). If the random number is smaller than the ratio of
  speciation and turnover, the lineage speciates, else it goes extinct:
\end{enumerate}

@@

The function \texttt{make.modifiers} allows to specifically change any
of these components by providing a different function for each part of
the algorithm. For example, we could modify the three functions above so
that the branch length is not dependent on the number of lineages, the
sampling is always the first species available and the extinction is
drawn from a normal distribution. Note that these function need a
specific syntax that is detailed in the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.

@@

This paper describes the first version of the \texttt{treats} package.
However, I intend, either personally or through collaborations to
continuously develop this package. For example future planned versions
will include: lineage specific simulations; abiotic events; time
dependent simulations and; a better integration with the
\texttt{dispRity} package. This will be done while keeping track change
(through \texttt{NEWS} file), continuous integration and unit testing.
Furthermore, I am planning to develop the GitHub page associated with
the package to encourage users to share their simulation templates and
build library from them to facilitate the work for future users.

@@

Thanks to Mark Puttick for inspiring to develop the package through
previous collaborations. Thanks to Andrew Beckerman, Ian Brennan,
Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston,
Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule,
Sophie Ryan and Gavin Thomas, for support and useful comments on late
stages of the development of this package. This work was funded by
UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research
Fellowship (URF R 180006 to Gavin Thomas).

>>>>

\title{\texttt{treats}: a modular \texttt{R} package for simulating
trees and traits.}
\author{Thomas Guillerme\(^1\)}
\date{2023-06-09}

@@

Comparing biological patterns is one of the key ways to understand
mechanisms in evolutionary biology. This leads to the development of
phylogenetic comparative methods as key methodologically driven topic in
ecology, evolution and palaeontology (Felsenstein 1985; Pennell and
Harmon 2013). These methods rely on comparing patterns in a phylogenetic
context to understand biological mechanisms or concepts (Harmon 2019).
These comparisons can be done between observed patterns under different
conditions or against null, neutral or baseline models (see Bausman
(2018) for distinctions) suggesting different processes or mechanisms.
For example different traits distribution for species with different
diets (Deepak, Gower, and Cooper 2023) or habitats (Pinto-Ledezma et al.
2017). Or by comparing some observed pattern to one simulated under null
or base conditions (Miller et al. 2022). In theory, workers can follow
the following research pipeline: thinking of a specific mechanism
(e.g.~mass extinction allowing the surviving species to acquire new
morphologies), collecting some data to test this mechanism (e.g.~some
traits of species across and extinction event) and then comparing these
patterns to some simulated under no specific conditions (e.g.~a null
model where the traits evolve randomly regardless of an extinction event
Puttick, Guillerme, and Wills (2020)). Workers might thus need to
simulate a great diversity of evolutionary scenarios to test their
specific question. To do so, we need statistical and software solutions
to simulate trees and data to generate many specific null models.

In practice, these evolutionary simulations can be done relatively
easily on computers using a birth-death process (Feller 1939; Stadler
2010; FitzJohn 2012). A birth-death process is a continuous time Markov
process that had been routinely implemented in \texttt{R} (R Core Team
2023) to simulate realistic phylogenies (e.g.~(Paradis and Schliep
2019), (FitzJohn 2012)). This general algorithm to generate phylogenetic
trees can be coupled with other Markov processes to also generate
traits, for example using a Brownian Motion process (BM; Cavalli-Sforza
and Edwards (1967)) or an Ornstein Uhlenbeck (OU; Lande (1976); see
Cooper et al. (2016) for a distinction between both). In \texttt{R},
this can be done with several already well used and well documented
packages. For example if you want to simulate diversity through time,
you can use \texttt{TreeSim} (Stadler 2011) to simulate diversity under
a set of specific parameters (e.g.~speciation and extinction) with some
events disrupting the simulations (e.g.~mass extinctions). You can even
improve on generating these patterns using \texttt{FossilSim}
(Barido-Sottani et al. 2019) to take into account fossilization
processes. You can also use \texttt{paleobuddy} (Rosario Petrucci,
Januario, and Quental 2022) or \texttt{paleotree} (Bapst 2012) to
generate palaeontology specific data. On the other hand, if you need to
simulate both diversity and traits through time, this can be done with
specific parameters in \texttt{RPANDA} (Morlon et al. 2016),
\texttt{diversitree} (FitzJohn 2012) or \texttt{PETER} (Puttick,
Guillerme, and Wills 2020) where the traits are generated stochasticaly
through time (given some process) during the birth-death process.

Although the packages mentioned above are excellent and routinely used
with fast and reliable algorithms and associated documentation, they are
all designed for specific tasks and don't allow much modification beyond
the input parameters designed by the authors. For example,
\texttt{TreeSim} can simulate a birth-death tree with some extinction
event but is not designed to simulated one with an extinction event that
leads to the birth-death process to be not diversity dependent anymore,
simulating a release in selection pressure after the extinction event
that leads to a different process dominating speciation. Or
\texttt{PETER} is not designed to simulate a complex set of traits (say
three correlated BM traits and two independent OU ones). This absence of
modularity has hampered the use of complex and question-driven
simulations, although I acknowledge this was not the primary aim of the
authors of the excellent packages mentioned above. This has led workers
to often develop their own tools to answer specific questions
(e.g.~Puttick, Guillerme, and Wills (2020)). Therefore, I propose
\texttt{treats} a modular \texttt{R} package to simulate both trees and
traits through time.

@@

These three steps are repeated until the tree reaches the desired age or
number of species (the algorithm's implementation is heavily inspired
and based on FitzJohn (2012)). If traits are simulated during the
process, a fourth step is added:

@@

In \texttt{treats}, these three or four steps are implemented as modular
functions that the user can easily change using an internal class of
objects called \texttt{"modifiers"} or \texttt{"traits"} (Fig.
\ref{Fig:workflow}). The simulation then outputs a tree (of class
\texttt{"phylo"} and a associated table of traits - \texttt{"matrix"})
that can be visualised using the \texttt{plot.treats} function. A third
class of object called \texttt{"events"} can be added to the simulations
to modify the entire simulation under certain conditions
(e.g.~simulating a mass extinction). Each element in the algorithm can
be modified by the user using the implemented functions
\texttt{make.bd.params} to set the birth-death parameters,
\texttt{make.traits} to set the trait(s), \texttt{make.modifiers} to set
the birth-death algorithm and \texttt{make.events} to eventually add one
or more events. Users can replace any step in the algorithm by their own
specific functions suiting their needs or, for less advanced users, by
using already implemented functions.

@@

To illustrate this we will look whether it is possible to detect changes
in disparity (i.e.~diversity of traits) in a subset of the data
published from Beck and Lee (2014) implemented in Guillerme (2018). This
dataset contains the ordinated traits for 50 mammalian species across
the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).

@@

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
\caption{Figure 2: Observed phylogeny (A) and disparity through time (B)
in the subset of Beck and Lee (2014)'s dataset. Is this change in
disparity related to the K-Pg mass extinction? Or, more generally, is it
even possible to detect potential changes of disparity due to a mass
extinction event?}
\end{figure}

@@

The first and simplest way is to simulate tree topologies that have
similar properties than the observed one. To do so, we need to use some
\textbf{speciation} parameter indicating the rate at which lineages
speciate (\emph{aka} ``birth'' or ``\(\lambda\)'') and an
\textbf{extinction} parameter indicating the rate at which they go
extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
be crudely calculated from a given tree by counting the number of
speciations and extinctions through time (but see better methods,
e.g.~\texttt{ape::birthdeath}; Lawton, May, and Raup (1995); Paradis and
Schliep (2019)). We also need a stopping rule for when to stop the
simulations (in our case when reaching 30 time units). This will produce
a single random tree using the input parameters (Fig. 3). Note that I
will not discuss the options here in great details. Much more
information can be found in the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.

@@

To answer our question, we also want to simulate an extinction event. To
do so, we can create two different \texttt{"events"} object with the
function \texttt{make.events}. The first one will simulate a random
extinction event after reaching half of the simulations (15 time units)
and then making three quarter (0.75) of the taxa go extinct. The second
one will simulate a random extinction but based on trait values: after
reaching time 15, all the species with positive trait values will go
extinct. Both scenarios thus illustrate two different types of mass
extinction (random and selective; Puttick, Guillerme, and Wills (2020)).

@@

From these results we can draw some preliminary conclusions: the
observed change in disparity is more likely due to a random mass
extinction rather than a selective one. This is of course a very crude
way of testing this, a more rigorous approach is needed to answer the
question: more and better quality data, and more thorough methods
(e.g.~using a rank envelope test Murrell (2018)).

@@

The following sections provides an overview of the three main functions
displayed in Figure 1 (\texttt{make.traits}, \texttt{make.modifiers} and
\texttt{make.events}).

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Brownian Motion process}
\NormalTok{my.bm.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, edge.length) \{}
    \KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ x0, }\DataTypeTok{sd =}\NormalTok{ edge.length)}
\NormalTok{\} }
\CommentTok{\#\# Creating the traits object}
\NormalTok{my.trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ my.bm.process)}
\end{Highlighting}
\end{Shaded}

Of course biological data can be much more complex and often
multivariate (Adams and Collyer 2019) requiring the users to develop
more complex function to cater their specific needs. The \texttt{treats}
package contains a list of pre-built processes that are
``ready-to-use'':

\begin{itemize}
\tightlist
\item
  \texttt{BM.process} and \texttt{OU.process}: generalised Brownian
  Motion and Ornstein-Uhlenbeck processes in any number of dimensions
  (including possible correlation);
\item
  \texttt{multi.peak.process}: a generalised Ornstein-Uhlenbeck process
  (uni- or multi-dimensional) which allows for multiple peaks;
\item
  \texttt{repulsion.process}: a unidimensional process generating trait
  values that don't overlap with previously generated trait values;
\item
  \texttt{discrete.process}: generates discrete trait values;
\item
  \texttt{no.process}: ignores the branch length (i.e.~a time
  independent process);
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  \textbf{Choosing whether that lineage goes extinct (becomes a tip) or
  speciates (becomes a node)} by drawing a random number between 0 and 1
  and comparing it to the ratio of speciation and turnover (speciation +
  extinction). If the random number is smaller than the ratio of
  speciation and turnover, the lineage speciates, else it goes extinct:
\end{enumerate}

@@

The function \texttt{make.modifiers} allows to specifically change any
of these components by providing a different function for each part of
the algorithm. For example, one can modify the three functions above so
that the branch length is not dependent on the number of lineages, the
sampling is always the first species available and the extinction is
drawn from a normal distribution. Note that these function need a
specific syntax that is detailed in the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.

@@

This paper describes the first version of the \texttt{treats} package.
However, I intend to continuously develop this package. For example
future planned versions will include: lineage specific simulations;
abiotic events; time dependent simulations and; a better integration
with the \texttt{dispRity} package. This will be done while keeping
track change (through \texttt{NEWS} file), continuous integration and
unit testing. Furthermore, I am planning to develop the GitHub page
associated with the package to encourage users to share their simulation
templates and build library from them to facilitate the work for future
users.

@@

Many thanks to Mark Puttick for inspiring the development the package
through previous collaborations. Thanks to Andrew Beckerman, Ian
Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex
Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James
Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on
late stages of the development of this package. This work was funded by
UKRI-NERC Grant NE/T000139/1.
